---
title: R graphics magic that you should never use
---
I've been making graphics in R for quite a while

I've spent years of my life trying to make nice plots with base R
graphics. I've learned a bunch of tricks along the way, so now I'm
going to show you everything I can remember.

## Data
I'm going to use data came from an earlier version of
[this website](http://peregrinossantiago.es/eng/post-peregrinacion/estadisticas/).
They're "statistics about pilgrims who have arrived in Santiago".

```r
pilgrams <- sqldf('select * from pilgram', dbname = 'pilgrams-db.sqlite')
```

I'm also going to use data from a
[wellbeing study](https://data.somervillema.gov/dataset/Results-of-2011-Wellbeing-Study/pjhx-dusc)
about Somerville, Massachusetts. There are a lot of missing values in this dataset, and I'm only
trying to show you how to make plots, so I'm just going to remove the missing values.

```r
wellbeing <- na.omit(read.csv('http://data.somervillema.gov/api/views/pjhx-dusc/rows.csv?accessType=DOWNLOAD'))
```

## One factor
Maybe we want count the marital statuses of Somerville residents.
This won't work.

    barplot(wellbeing$X24..What.is.your.marital.status.)

Instead, you need to give barplot a vector of heights. We can make one with `table`.

```{r barplot-
barplot(table(wellbeing$X24..What.is.your.marital.status.))
```

I can't read the bar labels, so let's make the bars horizontal.

```{r barplot-
barplot(table(wellbeing$X24..What.is.your.marital.status.), horiz = TRUE)
```

Oops. Now the labels are rotated the wrong way. The `las` parameter to `par` handles that.

```{r barplot-
barplot(table(wellbeing$X24..What.is.your.marital.status.), horiz = TRUE, las = 1)
```

And let's adjust the margin and bar labels so the text doesn't fall off.

```{r barplot-
levels(wellbeing$X24..What.is.your.marital.status.)[3] <- "Living with a partner\nbut not married"
par(mar = c(5, 12, 4, 2))
barplot(table(wellbeing$X24..What.is.your.marital.status.), horiz = TRUE, las = 1)
```

I also want to sort the bars so they're easier to read.

```{r barplot-
barplot(sort(table(wellbeing$X24..What.is.your.marital.status.), decreasing = TRUE), horiz = TRUE, las = 1)
```

The black border around the bars is noisy; let's remove it.

```{r barplot-
barplot(
  sort(table(wellbeing$X24..What.is.your.marital.status.), decreasing = TRUE),
  horiz = TRUE, las = 1, border = NA
)
```

Now the colors don't match. We could just make the bars black,

```{r barplot-
barplot(
  sort(table(wellbeing$X24..What.is.your.marital.status.), decreasing = TRUE),
  horiz = TRUE, las = 1, border = NA, col = 1
)
```

but that's a bit harsh. Instead, let's make everything grey.

```{r barplot-
barplot(
  sort(table(wellbeing$X24..What.is.your.marital.status.), decreasing = TRUE),
  horiz = TRUE, las = 1, border = NA, col = '#666666', fg = '#333333', col.axis = '#333333'
)
```

And add a title, of course

```{r barplot-
barplot(
  sort(table(wellbeing$X24..What.is.your.marital.status.), decreasing = TRUE),
  horiz = TRUE, las = 1, border = NA, col = '#666666', fg = '#333333', col.axis = '#333333',
  xlab = 'Number of residents',
  main = 'Marital statuses of Somerville residents'
)
```

The above call is boilerplate to me; in order to have any chance at making
any vaguely decent bar plot, you **minimally** need to adjust all of those parameters.
Summing that up, this is what I think the default barplot function should be.

```{r barplot-tom}
tom.barplot <- function(height, col = 'grey', border = NA, ...) {
  barplot(
    sort(table(height)), col = col, border = border,
    col = '#666666', fg = '#333333', col.axis = '#333333',
    ...
  )
}
```

## One factor and one numeric
A stripchart is sort of like a histogram, but it doesn't bin the observations.
Here is how people in Somerville are feeling, on a range of 1 to 10.

```{r stripchart-
stripchart(wellbeing$X1...First.of.all..how.happy.do.you.feel.right.now.., method = 'jitter')
```

We can also stack the observations

```{r stripchart-
stripchart(wellbeing$X1...First.of.all..how.happy.do.you.feel.right.now.., method = 'stack')
```

Let's change the scale to emphasize its limits.

```{r stripchart-
stripchart(wellbeing$X1...First.of.all..how.happy.do.you.feel.right.now..,
  method = 'stack', xlim = c(1:10)
)
```

The white boxes with borders are noisy; let's change them to filled boxes.
The default point type (`pch = 1`) doesn't support a fill color (`bg`),
so we have to switch to a different point type.

```{r stripchart-
stripchart(wellbeing$X1...First.of.all..how.happy.do.you.feel.right.now..,
  method = 'stack', xlim = c(1:10), col = NA, pch = 22, bg = 1
)
```

You can see all of the pch types here.

```{r stripchart-pch}
pchShow()
```

Now we could also color the points by a different color depending on the
gender.

```{r stripchart-
stripchart(wellbeing$X1...First.of.all..how.happy.do.you.feel.right.now..,
  method = 'stack', xlim = c(1:10), col = NA, pch = 22, bg = 3 + wellbeing$X15..Gender..1..female..2..male.
)
legend('topleft', c('Female', 'Male'), col = NA, pch = 22, pt.bg = 4:5
```

## par
All of these `col`, `pch`, `bg`, `fg`, `border`, &c. commands are graphical
parameters passed to `par`. You can set their global defaults with `par`,
but they won't necessarily apply to the same scope as when you pass them
within a function. For example, `bg` applies to the outer canvas's background.

```{r par-
par(bg = 'yellow')
plot(Total ~ id, data = pilgrams, type = 'l')
```

You might use this to set fonts across all of your plots.

```{r 
par(family = 'serif', font = 3, font.lab = 3, font.main = 2, font.sub = 3)
plot(Total ~ id, data = pilgrams, type = 'l', main = 'Pilgrams')
```

I tend to find the global namespace unpleasant, so I like to wrap that
in a function that closes the device afterwards, with `dev.off`.
(The settings only apply to the current device.)

    plot.something <- function(a, b, c, e, f) {
      par(family = 'serif', font = 3, font.lab = 3, font.main = 2, font.sub = 3)
      plot(a ~ b, col = c)
      lines(d, e)
      # ...
      dev.off()
    }

`mfrow` and `mfcol` allow you to make multiple plots in one device

```{r par-
par(mfrow = c(2,2))
plot(lm(Men ~ Women, data = pilgrams), mfrow = c(2,2))
```

`bty` lets you change the outer border of the plot.

```{r par-
plot(Total ~ id, data = pilgrams, type = 'l', bty = 'l')
```

Its options are quite hilarious; here is its entry in `?par`.

    ‘bty’ A character string which determined the type of ‘box’ which
         is drawn about plots.  If ‘bty’ is one of ‘"o"’ (the
         default), ‘"l"’, ‘"7"’, ‘"c"’, ‘"u"’, or ‘"]"’ the resulting
         box resembles the corresponding upper case letter.  A value
         of ‘"n"’ suppresses the box.

Most plotting commands call `plot.new`, which clears the plot by
default. Setting `new` to `TRUE` makes it not clear the plot, which
can help you build up more complicated plots.


## Scatterplots




## More scatterplots




    

```r
wellbeing$gender <- factor(wellbeing$X15..Gender..1..female..2..male., levels = c(1:2,NA), exclude = NULL)
levels(wellbeing$gender)[1:2] <- c('Female', 'Male')
table(wellbeing$gender)
```




## Conclusion
I hope that you see how much of a mess base R graphics is.
Learn from my mistakes! Don't torture yourself with base R graphics;
**use ggplot**!
